---
title: "EDA"
author: "David Nezelek"
date: "2022-12-05"
output: html_document
description: In this file we'll be performing the Exploratory data
             analysis part of the project
---


# Importing the necessary Libraries: 

```{r}
library(data.table)
library(anytime)
library(stringi)
library(stringr)
library(tidyverse)
```

# Reading the dataset:

```{r echo = False}
data_file_name = "C:/Users/nezel/Desktop/flights_miami_large.csv" 
data <- fread(data_file_name)
```

# Days in Advance Column
This column will show how many days before the actual flight was the data collected
We do this by subtracting the search date from the flight date.

``` {r echo = False}
data$daysInAdvance <- data$flightDate - data$searchDate
```

# Latest flights
Here we try to get the latest search for a particular flight by filtering by the daysInAdvance to the day right before the flight

```{r}
latestFlights <- data %>%
  group_by(legId) %>%
  filter(daysInAdvance == 1)
```

# Departure Hour Column:
This column will tell us which hour of the day is the flight departing. To do so: 
- we took the departure time in Epoch Seconds 
- we select the first 10 characters
- We convert those into numeric values
- we use anytime to parse the integers into a datetime structure
- Then we select the 2 characters which represent the hour

``` {r echo = False}
latestFlights$departureHour <- as.integer(
  substr(
    anytime(
      as.numeric(
        substr(
          latestFlights$segmentsDepartureTimeEpochSeconds, 1, 10
          )
        )
      ),12,13
    )
  )
```

# Arrival Hour Column
This column will tell us which hour of the day is the flight arriving. To do so:
- we took the arrival time in Epoch Seconds 
- we extract the first 10 digits after the '||' if they exist using regex
- We convert those into numeric values
- we use anytime to parse the integers into a datetime structure
- Then we select the 2 characters which represent the hour

``` {r echo = False}
latestFlights$arrivalHour <- as.integer(
  substr(
    anytime(
      as.numeric(
        stri_extract(
          latestFlights$segmentsArrivalTimeEpochSeconds, regex="[0-9]{10}(?!\\|)"
                     )
        )
      )
    ,12,13
    )
  )
```

# Layover Duration in Hours
This column gives the total minutes between connecting flights. For nonstop flights, this value will equal 0.
-We first convert the travelDuration variable to a number of seconds. 
-We then find the total time in the air, encoded in the segmentsDurationInSeconds variable. 
-The difference between these two calculated values is the layover duration in seconds. 
-Finally, we convert to hours by dividing by 3600.

``` {r echo = False}
hours_converted <- 3600*as.numeric(stri_extract(latestFlights$travelDuration,regex="(?<=PT)[0-9]{1,2}(?=H)"))
minutes_converted <- 60*as.numeric(stri_extract(latestFlights$travelDuration,regex="(?<=H)[0-9]{1,2}(?=M)"))
total_trip_duration_in_seconds <- hours_converted + minutes_converted

time_in_air <- str_split(latestFlights$segmentsDurationInSeconds,"\\|\\|") 
total_time_in_air <- sapply(lapply(time_in_air, as.numeric),sum)

layover_duration_in_seconds <- total_trip_duration_in_seconds - total_time_in_air

latestFlights$layoverDurationInHours <- layover_duration_in_seconds/3600
```

# First Airline
Here we extract the first word from the given segmentsAirlineName variable. This classifies each flight, whether nonstop or connecting, by its first (or only) airline.
``` {r echo = False}
latestFlights$firstAirline <- stri_extract(latestFlights$segmentsAirlineName, regex="\\w+")
```


```{r}
ggplot(data = latestFlights, aes(x = seatsRemaining, y = totalFare)) + geom_point() + ggtitle("Fares by Seats Remaining")
ggplot(data = latestFlights, aes(x = seatsRemaining, y = totalFare, color = firstAirline)) + geom_point() + ggtitle("Fares by Seats Remaining, Colored by Airline")
ggplot(data = latestFlights, aes(x = totalFare)) + geom_histogram() + ggtitle("Counts of Flights by Total Fare")
ggplot(data = latestFlights, aes(x = seatsRemaining)) + geom_histogram() + ggtitle("Counts of Flights by Seats Remaining") + scale_x_continuous(breaks = seq(-1, 11, 1), lim = c(-1, 11))
ggplot(data = latestFlights, aes(x = seatsRemaining, fill = firstAirline)) + geom_histogram() + ggtitle("Counts of Flights by Seats Remaining, Colored by Airline") + scale_x_continuous(breaks = seq(-1, 11, 1), lim = c(-1, 11))
ggplot(data = latestFlights, aes(x = layoverDurationInHours, y = totalFare, color = destinationAirport)) + geom_point() + ggtitle("Fares by Length of Layover, Colored by Destination")
ggplot(data = latestFlights, aes(x = arrivalHour, y = totalFare, color = destinationAirport)) + geom_point() + ggtitle("Fares by Hour of Arrival, Colored by Destination")
ggplot(data = latestFlights, aes(x = departureHour, y = totalFare, color = destinationAirport)) + geom_point() + ggtitle("Fares by Hour of Departure, Colored by Destination")
```
